Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-08 manuscript/manuscript.tex
changes to disco

Since the seminal analysis of \textcite{Maynard_Smith1974-lc} demonstrating
that linked neutral diversity is reduced as an advantageous polymorphism arises
and sweeps to fixation, over four decades of theoretical and empirical research
has enriched our understanding of linked selection.  One under-used approach to
understand the genome-wide effects of selection on standing variation, e.g.
selection on an infinitesimal polygenic trait, stems from an early quantitative
genetic model of linked selection \parencite{Robertson1961-ho} and its later
developments
(\cite{Santiago1995-hx,Santiago1998-bs,Wray1990-zf,J_A_Woolliams_N_R_Wray_R_Thompson2008-qo};
see also \cite{Barton2000-zg} for a comparison of these models with classic
hitchhiking models). Implicit in these models is that autocovariance between
allele frequency change is created when there is heritable fitness variation in
the population, a signal that may be readily detected from temporal genomic
data \parencite{Buffalo2019-io}.  Depending on how many loci affect fitness,
such a strong effect of linked selection may not be differentiable from genetic
drift using only single contemporary population samples or looking at at
temporal allele frequency change at each locus in isolation.  In this way,
averaging summaries of temporal data allows us to sidestep the key problem of
detecting selection from standing variation: that the genomic footprint leaves
too soft of a signature to differentiate from a background of genetic drift. In
fact we find that the temporal covariance signal is detectable even in the most
extremely difficult to detect soft sweep case: polygenic selection on highly
polygenic traits \parencite{Buffalo2019-io}.

@@

Our analyses reveal that a sizable proportion of allele frequency change in
these populations is due to the (indirect) action of selection. Capitalizing on
replicated designs, we characterized the extent to which convergent selection
pressures lead to parallel changes in allele frequencies across replicate
populations, and found that a reasonable proportion of the response is shared
across short timescales. These likely represent substantial under-estimates of
the contribution of linked selection because \vb{the studies we have reanalyzed
do not sequence the population each generation, preventing us from including
the effects of stronger correlations between adjacent generations.
Furthermore, estimation methods are intentionally conservative, further
leading our estimates to be closer to a lower bound of the effects of linked
selection than a measure of central tendency.}

@@

Our primary focus here has been on evolution in laboratory populations. It is
unclear whether we should expect a similar impact of selection in natural
populations. In some of these experiments, selection pressures may have been
stronger or more sustained that in natural populations
\vb{\parencite{Hendry1999-zu,Hairston2005-ga}}. Conversely, these lab
populations were maintained at very small population sizes, small effective
population sizes 300, 450, and 45 for the \textcite{Kelly2019-dc}, and
\textcite{Castro2019-uk} studies respectively, which will amplify the role of
genetic drift. The advantage of lab experiments is that they are closed
populations, in natural populations temporal covariance could also arise from
the systematic migration of alleles from differentiated populations.
\vb{Adapting} these methods to natural populations will require either
populations that are reasonably closed to migration, or for the effect of
migration to be accounted for possibly either by knowledge of allele
frequencies in source populations or the identification of migrant individuals.

While it challenging to apply temporal methods to natural populations there is
a lot of promise for these approaches
\parencite{Bergland2014-ij,Machado2018-cs}. Efforts to estimate the quantify
the impact of linked selection in have found over obligately sexual organisms
\vb{0\% to 89\%} reduction in genome-wide diversity over long time periods is
driven by linked selection
\parencite{McVicker2009-ax,Elyashiv2016-vt,Corbett-Detig2015-gt,Coop2016-gx,Comeron2014-nh}
Thus there is reason to be hopeful that we could detect this from temporal
data, which would help to resolve the timescales that linked selection act
over. In our reanalysis of the \textcite{Barghi2019-qy} study, we find evidence
of complex linked selection dynamics, with selection pressures flipping over
time due to either environmental change, the breakup of epistatic combinations
or advantageous haplotypes. Such patterns would be completely obscured in
samples only from contemporary populations, thus we can hope to have a much
richer picture of linked selection as temporal sequencing becomes more common,
\vb{perhaps allowing us to observe the effects of ecological dynamics in
genomic data \parencite{Hairston2005-ga}.}

>>>>

Since the seminal analysis of \textcite{Maynard_Smith1974-lc} demonstrating
that linked neutral diversity is reduced as an advantageous polymorphism arises
and sweeps to fixation, over four decades of theoretical and empirical research
has enriched our understanding of linked selection.  One under-used approach to
understand the genome-wide effects of selection on standing variation, e.g.
selection on an infinitesimal polygenic trait, stems from an early quantitative
genetic model of linked selection \parencite{Robertson1961-ho} and its later
developments
(\cite{Santiago1995-hx,Santiago1998-bs,Wray1990-zf,J_A_Woolliams_N_R_Wray_R_Thompson2008-qo};
\gc{THE WRAY PAPER DATE IS WRONG}
see also \cite{Barton2000-zg} for a comparison of these models with classic
hitchhiking models). Implicit in these models is that autocovariance between
allele frequency change is created when there is heritable fitness variation in
the population, a signal that may be readily detected from temporal genomic
data \parencite{Buffalo2019-io}.  Depending on how many loci affect fitness,
such a strong effect of linked selection may not be differentiable from genetic
drift using only single contemporary population samples or looking at at
temporal allele frequency change at each locus in isolation.  In this way,
averaging summaries of temporal data allows us to sidestep the key problem of
detecting selection from standing variation: that the genomic footprint leaves
too soft of a signature to differentiate from a background of genetic drift. In
fact we find that the temporal covariance signal is detectable even in the most
extremely difficult to detect soft sweep case: polygenic selection on highly
polygenic traits \parencite{Buffalo2019-io}.

@@

Our analyses reveal that a sizable proportion of allele frequency change in
these populations is due to the (indirect) action of selection. Capitalizing on
replicated designs, we characterized the extent to which convergent selection
pressures lead to parallel changes in allele frequencies across replicate
populations, and found that a reasonable proportion of the response is shared
across short timescales. These likely represent substantial under-estimates of
the contribution of linked selection because \vb{the studies we have reanalyzed
do not sequence the population each generation, preventing us from including
the effects of stronger correlations between adjacent generations.
Furthermore, \gc{our} estimation methods are intentionally
conservative,\gc{for example they exclude the contribution of selection that
  does not persist across generations and selection that reverses sign
  thus they show be seen as a storng lower bound of the effects of 
selection.}

@@

Our primary focus here has been on evolution in laboratory populations. It is
unclear whether we should expect a similar impact of selection in natural
populations. In some of these experiments, selection pressures may have been
stronger or more sustained that in natural populations
\vb{\parencite{Hendry1999-zu,Hairston2005-ga}}. Conversely, these lab
populations were maintained at very small population sizes, small effective
population sizes 300, 450, and 45 for the \textcite{Kelly2019-dc}, and
\textcite{Castro2019-uk} \gc{MISSING REF??} studies respectively, which will amplify the role of
genetic drift. The advantage of lab experiments is that they are closed
populations, in natural populations temporal covariance could also arise from
the systematic migration of alleles from differentiated populations.
\vb{Adapting} these methods to natural populations will require either
populations that are reasonably closed to migration, or for the effect of
migration to be accounted for possibly either by knowledge of allele
frequencies in source populations or the identification of migrant individuals.

While it challenging to apply temporal methods to natural populations there is
a lot of promise for these approaches
\parencite{Bergland2014-ij,Machado2018-cs}. Efforts to estimate the quantify
the impact of linked selection in have found over obligately sexual organisms
a \vb{0\% to 89\%} reduction in genome-wide diversity over long time periods is
driven by linked selection
\parencite{McVicker2009-ax,Elyashiv2016-vt,Corbett-Detig2015-gt,Coop2016-gx,Comeron2014-nh}
\gc{Thus linked selection makes a sizeable contribution to long-term allele
frequency change in some species,} and there is reason to be hopeful that we could detect this from temporal
data, which would help to resolve the timescales that linked selection act
over. In our reanalysis of the \textcite{Barghi2019-qy} study, we find evidence
of complex linked selection dynamics, with selection pressures flipping over
time due to either environmental change, the breakup of epistatic combinations
or advantageous haplotypes. Such patterns would be completely obscured in
samples only from contemporary populations. \gc{Thus,} we can hope to have a much
richer picture of the\gc{ impact of selection} as temporal sequencing becomes more common,
\gc{allowing} us to observe the effects of ecological dynamics in
genomic data \parencite{Hairston2005-ga}.
